---
title: "Peer-graded Assignment: Course Project 1"
author: "Giovanna Gomez"
date: "11/13/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

setwd("~/Dropbox/Education/JHU/Data Sience Specialization/5. Reproducible Research/wd_for_rr")
```

## Loading and preprocessing the data

```{r}
# Load the data.

activity <- read.csv("activity.csv", na.strings = "NA")

# Process/transform the data (if necessary) into a format suitable for your analysis.

str(activity)

library(lubridate)

activity$date <- ymd(activity$date)

head(activity); tail(activity)
```

## What is mean total number of steps taken per day?

```{r, fig.align = 'center'}
attach(activity)

# Calculate the total number of steps taken per day.

total_steps <- tapply(steps, date, sum, na.rm = TRUE)

total_steps

# Make a histogram of the total number of steps taken each day.

hist(total_steps, main = "Histogram of the Total Number of Steps", xlab = "Total number of steps taken each day", las = 1)

# Calculate and report the mean and the median of the total number of steps taken per day.

summary(total_steps)
```

## What is the average daily activity pattern?

```{r, fig.align = 'center'}
# Make a time series plot of the 5-minute interval and the average number of steps taken.

average_steps <- round(tapply(steps, interval, mean, na.rm = TRUE), digits = 2)

interval.f <- activity$interval

interval.f <- unique(interval.f)

plot(interval.f, average_steps, type = "l", main = "Time Series Plot", xlab = "5-minute interval", ylab = "Average number of steps", las = 1)

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

max(average_steps)

both <- data.frame(interval.f, average_steps)

both[which.max(both$average_steps), ]
```

## Imputing missing values

```{r,fig.align = 'center'}
# Calculate and report the total number of missing values in the dataset.

sum(is.na(activity))

# Devise a strategy for filling in all of the missing values in the dataset.

sum(is.na(steps))

detach(activity)

activity[is.na(activity)] <- round(average_steps, 0)

# Create a new dataset that is equal to the original dataset but with the missing data filled in.

activity2 <- activity

str(activity2)

head(activity2); tail(activity2)

# Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

attach(activity2)

total_steps2 <- tapply(steps, date, sum)

hist(total_steps2, main = "Histogram of the New Total Number of Steps", xlab = "New Total number of steps taken each day", las = 1)

summary(total_steps2)

detach(activity2)
```

**Do these values differ from the estimates from the first part of the assignment?**

The mean increased a little bit (9354 vs 10770) because I used the average number of steps in order to replace the missing data; on the other hand, the median almost remained the same (10400 vs 10760).

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**

The impact was noticeable for the mean.

## Are there differences in activity patterns between weekdays and weekends?

```{r, fig.align = 'center'}
# Create a new factor variable in the dataset  with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

activity2$date <- weekdays(activity2$date)

activity2$date2[activity2$date == "Monday"] <- "weekday"

activity2$date2[activity2$date == "Tuesday"] <- "weekday"

activity2$date2[activity2$date == "Wednesday"] <- "weekday"

activity2$date2[activity2$date == "Thursday"] <- "weekday"

activity2$date2[activity2$date == "Friday"] <- "weekday"

activity2$date2[activity2$date == "Saturday"] <- "weekend"
    
activity2$date2[activity2$date == "Sunday"] <- "weekend"

str(activity2$date2)

activity2$date2 <- factor(activity2$date2)

str(activity2$date2)

# Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday or weekend days.

weekday <- activity2[activity2$date2 == "weekday", ]

weekday_average_steps <- round(tapply(weekday$steps, weekday$interval, mean), digits = 2)

weekend <- activity2[activity2$date2 == "weekend", ]

weekend_average_steps <- round(tapply(weekend$steps, weekend$interval, mean), digits = 2)

par(mfrow = c(2,1), oma = c(1, 1, 2, 1))

plot(interval.f, weekday_average_steps, type = "l", xlab = "5-minute interval", ylab = "Average number of steps", main = "weekdays", las = 1)

plot(interval.f, weekend_average_steps, type = "l", xlab = "5-minute interval", ylab = "Average number of steps", main = "weekends", las = 1)

mtext("Time Series Plot by Weekdays and Weekends", outer = TRUE, cex = 1.5)
```

**Conclusion:** Yes, there are some differences in activity patterns between weekdays and weekends.